In-silico molecular enrichment and clearance of the human intracranial space

### 2025-06-03 paper/main.tex
Updates in response to reviewer's comments on enlarged PVS.

\begin{abstract}
  The mechanisms of intracranial solute transport are fundamental to human brain health, with alterations often linked to disease and functional impairment, and with distinct opportunities for personalized diagnostics and treatment. However, our understanding of these mechanisms and their interplay remains incomplete, in part due to the complexity of integrating insights across scales, between species and from different modalities. Here, we combine mixed-dimensional modelling, multi-modal magnetic resonance images, and high performance computing to construct and explore a high-fidelity in-silico model of human intracranial molecular enrichment.  This model predicts the temporo-spatial spreading of a solute within an image-derived geometric representation of the subarachnoid space, ventricular system and brain parenchyma, including networks of surface perivascular spaces (PVSs). Our findings highlight the significant impact of cerebrospinal fluid (CSF) production and intracranial pulsatility on molecular enrichment following intrathecal tracer injection. We demonstrate that low-frequency vasomotion induces moderate CSF flow in surface PVS networks which substantially enhances tracer enrichment, and that impaired enrichment is a direct natural consequence of enlarged PVSs. This openly available technology platform thus provides an opportunity for integrating separate observations on diffusion in neuropil, vascular dynamics, intracranial pulsatility, CSF production, and efflux, and for exploring drug delivery and clearance in the human brain.
\end{abstract}
\begin{document}

@@

\subsection*{Enlarged PVSs delay periarterial and intracranial molecular enrichment}

Enlarged PVSs are associated with impaired periarterial tracer
transport and more generally with cognitive decline as well as
certain subtypes of
dementia~\cite{bown2022physiology,eide2024functional}. A key
question is thus whether enlarged PVSs alone leads to delayed perivascular
and intracranial tracer enrichment. To address this question from a
fluid and transport dynamics perspective, we ask our in-silico model
to predict the integrated effect of perivascular dilation on PVS flow
velocities and tracer enrichment patterns (\Cref{fig:enlargement}A).

Again we compute CSF production-, cardiac-, and vasomotion-induced net PVS flow velocities, but now in a perivascular geometry with an increased outer radius ($R^{\rm dilated}_2 = 3 R_1$) compared to the previous ($R^{\rm control}_2 = 2 R_1$), which corresponds to a $2.67 \times$ increase in PVS cross-section area.
%% \begin{align*}
%%   A_1 = \pi (R_2^2 - R_1^2) = \pi (2^2 R_1^2 - R_1^2) = 3 \pi R_1^2 \\
%%   A_2 = \pi (\hat{R}_2^2 - R_1^2) = \pi (3^2 R_1^2 - R_1^2) = 8 \pi R_1^2 \\
%%   A_2 / A_1 = 8/3 = 2.76
%% \end{align*}
The enlargement yields notable and non-uniform changes in the flow
dynamics. On the one hand, the pressure-induced flow velocities
increase from a mean of 0.08 \textmu m/s to 0.33 \textmu m/s due to
the reduced effective resistance of the PVS. However, the peristaltic
pumping becomes considerably less effective: the cardiac and
vasomotion-induced net CSF flow velocities drop from a mean of 0.92
\textmu m/s to 0.17 \textmu m/s and from 13.05 \textmu m/s to 2.33
\textmu m/s, respectively (\Cref{fig:enlargement}B--D). In total, the
net CSF velocity reduces, which in turn alters the tracer enrichment
within the dilated PVS. Notably, we observe later tracer arrival in
all MCA segments, with up to 96 min later arrival in the
MCA2s for the high PVS flow scenario (\Cref{fig:enlargement}E, F). For
the baseline model, which omits the vasomotion-induced net PVS flow,
the effect is less pronounced in the MCA but clearly persists and is
evident in the MCA2s (\Cref{fig:enlargement}F). Finally, we note that
the larger volume of the dilated PVS results in a greater total
accumulation of tracer in the PVS from 3--24 hours and reduced
enrichment of the parenchyma at 6 and 12 h, while the mean PVS
concentration is initially lower, but later exceeds the baseline model (\Cref{fig:enlargement}G,H).
\begin{figure}
    \centering
    \includegraphics[width=1 \textwidth]{figures/figure5.png}
\caption{\textbf{PVS enlargement reduces net PVS flow and delays tracer transport}
A) Modelling dilated PVS by extending the outer PVS boundaries;
B) mean and max PVS flow velocities with normal/control and dilated PVS;
C--D) 3D rendering of vasomotion-induced PVS flow in the normal (C) and dilated (D) periarterial networks; 
E) 2D slices showing zoom-in to the region surrounding the MCA-R after 1, 2, 3, and 4\,h with normal (lower) and dilated (upper) PVS. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF)); 
F) First-time arrival for different branches of the arterial tree for the baseline model (with pressure-driven and cardiac-induced net PVS flow) and the high PVS flow model (also including vasomotion-induced net PVS flow) for normal and dilated PVS;
G) total tracer content in the CSF, parenchyma and arterial and venous PVS at 1,3,6,12 and 24\,h for the four models from F;
H) mean concentration in the CSF, parenchyma and arterial and venous PVS over time for the four models from F.
}
    \label{fig:enlargement}
\end{figure}

@@

%% Summary of findings
We have presented a high-fidelity in-silico model of molecular
enrichment and clearance in human intracranial spaces, enabling
tailored predictions of the influence of CSF space and (peri)vascular
morphology, physiological factors such as cardiac and respiratory
pulsatility, as well as of pathological conditions such as enlarged
PVSs. A key observation is that the balance between CSF production and
intracranial pulsatility significantly shapes the large-scale features
of molecular enrichment, with e.g.~ventricular tracer reflux after
intrathecal injection in the absence of CSF production, and preference
towards anterior brain regions with reduced pulsatility. Indeed, the
simulated distribution patterns cover a substantial range of the
individual variations observed clinically \cite{ringstad2018brain}. In
terms of perivascular pathways, we find that even moderate CSF flow on
the order of 0.1--1.0 \textmu m/s in the surface PVSs results in
earlier tracer enrichment around major cerebral arteries. This effect
is more pronounced -- with up to a three-hour difference in time of
arrival -- with mean PVS flow rates of approximately 10 \textmu m/s. Our
models of CSF flow in perivascular networks, based on first principles
and asymptotic analysis, predict net flow of such magnitudes and both
antegrade and retrograde flow within the perivascular network,
occurring with or without pressure differences at the network
ends. The PVS may thus function as highways facilitating rapid
transport, even in the absence of structural barriers, though a sharp
concentration difference between the PVS and surrounding tissues is
unlikely without such barriers. Finally, dilated PVS will cause a
substantial reduction in cardiac- and vasomotion-driven flow
velocities, which in turn impedes perivascular transport.

@@

    & dilated PVS & $R_2 = 3 R_1$ & $-$ & $-$ & $-$ & $-$ & $\tilde{u}_a^p + \tilde{u}_a^c$  & $-$ \\ 

>>>>

\begin{abstract}
  The mechanisms of intracranial solute transport are fundamental to human brain health, with alterations often linked to disease and functional impairment, and with distinct opportunities for personalized diagnostics and treatment. However, our understanding of these mechanisms and their interplay remains incomplete, in part due to the complexity of integrating insights across scales, between species and from different modalities. Here, we combine mixed-dimensional modelling, multi-modal magnetic resonance images, and high performance computing to construct and explore a high-fidelity in-silico model of human intracranial molecular enrichment.  This model predicts the temporo-spatial spreading of a solute within an image-derived geometric representation of the subarachnoid space, ventricular system and brain parenchyma, including networks of surface perivascular spaces (PVSs). Our findings highlight the significant impact of cerebrospinal fluid (CSF) production and intracranial pulsatility on molecular enrichment following intrathecal tracer injection. We demonstrate that low-frequency vasomotion induces moderate CSF flow in surface PVS networks which substantially enhances tracer enrichment, and that impaired enrichment is a direct natural consequence of enlarged surface PVSs. This openly available technology platform thus provides an opportunity for integrating separate observations on diffusion in neuropil, vascular dynamics, intracranial pulsatility, CSF production, and efflux, and for exploring drug delivery and clearance in the human brain.
\end{abstract}
\begin{document}

@@

\subsection*{Enlarged PVSs in the SAS delay periarterial and intracranial molecular enrichment}

Idiopathic normal pressure hydrocephalus (iNPH), a dementia subtype,
is associated with enlarged PVS in the SAS and impaired periarterial
and parenchymal tracer transport~\cite{eide2024functional}. More
generally, enlarged PVSs are associated with a range of pathophysiogical
conditions and cognitive decline~\cite{bown2022physiology}. A key
question is thus whether enlarged PVSs alone leads to delayed
perivascular and intracranial tracer enrichment. To address this
question from a fluid and transport dynamics perspective, we ask our
in-silico model to predict the integrated effect of perivascular
dilation on PVS flow velocities and tracer enrichment patterns
(\Cref{fig:enlargement}A).

Again we compute CSF production-, cardiac-, and vasomotion-induced net PVS flow velocities, but now in a perivascular geometry with an increased outer radius ($R^{\rm dilated}_2 = 3 R_1$) compared to the previous ($R^{\rm control}_2 = 2 R_1$), which corresponds to a $2.67 \times$ increase in PVS cross-section area.
%% \begin{align*}
%%   A_1 = \pi (R_2^2 - R_1^2) = \pi (2^2 R_1^2 - R_1^2) = 3 \pi R_1^2 \\
%%   A_2 = \pi (\hat{R}_2^2 - R_1^2) = \pi (3^2 R_1^2 - R_1^2) = 8 \pi R_1^2 \\
%%   A_2 / A_1 = 8/3 = 2.76
%% \end{align*}
The enlargement yields notable and non-uniform changes in the flow
dynamics. On the one hand, the pressure-induced flow velocities
increase from a mean of 0.08 \textmu m/s to 0.33 \textmu m/s due to
the reduced effective resistance of the PVS. However, the peristaltic
pumping becomes considerably less effective: the cardiac and
vasomotion-induced net CSF flow velocities drop from a mean of 0.92
\textmu m/s to 0.17 \textmu m/s and from 13.05 \textmu m/s to 2.33
\textmu m/s, respectively (\Cref{fig:enlargement}B--D). In total, the
net CSF velocity reduces, which in turn alters the tracer enrichment
within the dilated PVS. Notably, we observe later tracer arrival in
all MCA segments, with up to 96 min later arrival in the
MCA2s for the high PVS flow scenario (\Cref{fig:enlargement}E, F). For
the baseline model, which omits the vasomotion-induced net PVS flow,
the effect is less pronounced in the MCA but clearly persists and is
evident in the MCA2s (\Cref{fig:enlargement}F). Finally, we note that
the larger volume of the dilated PVS results in a greater total
accumulation of tracer in the PVS from 3--24 hours and reduced
enrichment of the parenchyma at 6 and 12 h, while the mean PVS
concentration is initially lower, but later exceeds the baseline model (\Cref{fig:enlargement}G,H).
\begin{figure}
    \centering
    \includegraphics[width=1 \textwidth]{figures/figure5.png}
\caption{\textbf{PVS enlargement in the SAS reduces net PVS flow and delays tracer transport}
A) Modelling dilated PVS in the SAS by extending the outer PVS boundaries;
B) mean and max PVS flow velocities with normal/control and dilated PVS;
C--D) 3D rendering of vasomotion-induced PVS flow in the normal (C) and dilated (D) periarterial networks; 
E) 2D slices showing zoom-in to the region surrounding the MCA-R after 1, 2, 3, and 4\,h with normal (lower) and dilated (upper) PVS. Innermost circle indicates the artery (blue), the surrounding annulus is the PVS (color represents concentration value), the surroundings are parenchyma and/or CSF spaces, separated by the pia (blue: towards the parenchyma, pink: towards the CSF)); 
F) First-time arrival for different branches of the arterial tree for the baseline model (with pressure-driven and cardiac-induced net PVS flow) and the high PVS flow model (also including vasomotion-induced net PVS flow) for normal and dilated PVS;
G) total tracer content in the CSF, parenchyma and arterial and venous PVS at 1,3,6,12 and 24\,h for the four models from F;
H) mean concentration in the CSF, parenchyma and arterial and venous PVS over time for the four models from F.
}
    \label{fig:enlargement}
\end{figure}

@@

%% Summary of findings
We have presented a high-fidelity in-silico model of molecular
enrichment and clearance in human intracranial spaces, enabling
tailored predictions of the influence of CSF space and (peri)vascular
morphology, physiological factors such as cardiac and respiratory
pulsatility, as well as of pathological conditions such as enlarged
PVSs in the SAS. A key observation is that the balance between CSF
production and intracranial pulsatility significantly shapes the
large-scale features of molecular enrichment, with e.g.~ventricular
tracer reflux after intrathecal injection in the absence of CSF
production, and preference towards anterior brain regions with reduced
pulsatility. Indeed, the simulated distribution patterns cover a
substantial range of the individual variations observed clinically
\cite{ringstad2018brain}. In terms of perivascular pathways, we find
that even moderate CSF flow on the order of 0.1--1.0 \textmu m/s in
the surface PVSs results in earlier tracer enrichment around major
cerebral arteries. This effect is more pronounced -- with up to a
three-hour difference in time of arrival -- with mean PVS flow rates
of approximately 10 \textmu m/s. Our models of CSF flow in
perivascular networks, based on first principles and asymptotic
analysis, predict net flow of such magnitudes and both antegrade and
retrograde flow within the perivascular network, occurring with or
without pressure differences at the network ends. The PVS may thus
function as highways facilitating rapid transport, even in the absence
of structural barriers, though a sharp concentration difference
between the PVS and surrounding tissues is unlikely without such
barriers. Finally, dilated PVS in the SAS will cause a substantial reduction in
cardiac- and vasomotion-driven flow velocities, which in turn impedes
perivascular transport.

@@

    & dilated PVS in the SAS & $R_2 = 3 R_1$ & $-$ & $-$ & $-$ & $-$ & $\tilde{u}_a^p + \tilde{u}_a^c$  & $-$ \\ 
